---
title: "Assignment 5: Mixed models"
author: "Marton Kovacs / Zoltan Kekecs"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Introduction

This assignment is related to the previous lab assignment concerning perioperative pain and its psychological and hormonal predictors. Just like previously, you will set up linear model to predict postoperative pain after wisdom tooth surgery, but this time you will have to also take into account the fact that there is clustering in the data. 

Your research paper on the effect of the psychological and hormonal predictors of postoperative pain was so successful, that you managed to secure research funding for a multi-site replication study. Here your collaborators collect data in the same way you did in the original study at 20 different hospital sites. The goal of the study is to increase the generalizability of your findings. You would like to assess the model coefficients and the overall predictive efficiency of the predictors in your model.

As a reminder, here is the protocol for data collection: “You have collected data from adults who were scheduled to undergo surgical extraction of the third mandibular molar (wisdom tooth surgery). Patients filled out a form in the waiting room before their surgery. The form contained questions about their sex, age, and weight, and psychological questionnaires assessing anxiety, pain catastrophizing, and mindfulness (see descriptions below). You also got blood samples and saliva samples from participants in the waiting room 5 minutes before their operations to determine the serum (a component of the blood) and salivary cortisol levels of participants. Participants were contacted 5 hours after the surgery to see how much pain they were experiencing. The __level of pain__ at that moment was recorded using a numerical rating scale using a __scale of 0 to 10__, where 0 means “no pain” and 10 means “worst pain I can imagine”. 

__The State Trait Anxiety Inventory:__ T measures trait anxiety on a scale of 20 to 80, higher scores mean higher anxiety. Anxiety has been found in many studies to positively correlate with the level of pain experienced. This is __variable STAI_trait__ in the dataset. 

__The Pain Catastrophizing Scale__ measures the extent of pain catastrophizing, which is characterized by a tendency to magnify the threat value of a pain stimulus and to feel helpless in the presence of pain, as well as by a relative inability to prevent or inhibit pain-related thoughts in anticipation of, during, or following a painful event. The total score on this scale ranges from 0 to 52, higher scores mean higher catastrophizing. Pain catastrophizing is one of the well-established predictors of clinical pain. This is __variable pain_cat__ in the dataset.

__The Mindful Attention Awareness Scale (MAAS)__ measures dispositional mindfulness, which may be described as a tendency to turn attention to present-moment experiences in an open, non-judgemental way. The MAAS total score ranges from 1 to 6 (an average of the item scores), with higher scores representing higher dispositional mindfulness. Trait mindfulness has been theorized to serve as a protective factor against pain, as the individual would be more objective about their pain experience and tend to associate less discomfort, despair, and hopelessness to the pain-related sensations. This is __variable mindfulness__ in the dataset.

__Cortisol__ is a stress hormone associated with acute and chronic stress. Cortisol levels are thought to be positively associated with pain experience. Cortisol can be __measured from both blood and the saliva__, although, serum cortisol is often regarded in medical research as more reliably related to stress (serum is a component of the blood plasma). These are __variables cortisol_serum__, and __cortisol_saliva__ in the dataset.”

# Datasets

You will need two datasets for this assignment, datafile A and B. You can load them from the 'data/' folder.

# Task

First, fit a linear mixed model to estimate postoperative pain on datafile A. You should use the same fixed effect predictors as you used in your final model in the 3 assignment. (If you did not do that assignment, use the following predictors: age, sex, STAI, pain catastrophizing, mindfulness, and serum cortisol.) Importantly, the model has to account for the clustering of the data in different hospital sites. We have no reason to assume that the effects of the different predictors would be different in the different hospitals, so fit a random intercept model including the random intercept of hospital-ID. Once the model is built, note the model coefficients and the confidence intervals of the coefficients for all fixed effect predictors, and compare them to the ones obtained in the 3 assignment. 

Also, compute the variance explained by the fixed effect predictors using marginal R^2^, and the variance explained by the fixed and random effect terms combined using conditional R^2^. Now use the model coefficients obtained on data file A to predict pain in datafile B.

__IMPORTANT:__ Do not fit the regression models on data file B (don’t re-train your models), just use the regression equation you derived based on datafile A. These regression equations should be applied on the new data (datafile B), to predict pain.

Now compute the variance explained by the model on datafile B. You can do this by using the formula: __1 - (RSS / TSS) = R^2__. Compare this R^2^ to the marginal and conditional R^2^ values computed for the model on datafile A. 

# What to report

Report the model coefficients and the confidence intervals of the coefficients for each fixed effect predictor obtained on data file A in a table. 

Report the variance components for the fixed effects, the random intercept, and the residuals (from the model on data file A). Also report the marginal R^2^ and the conditional R^2^ squared obtained from the model on data file A, and the observed R^2^ of this model for data file B.

# What to discuss

Compare the model coefficients and the confidence intervals observed in this assignment and the assignment for Lab 2 and discuss what you think the differences or similarities mean.

# Solution

## Read the data

Read the datasets used in this assignment. Pay attention to the extensions of the datafiles.

```{r}

#install.packages("haven") 

library(tidyverse)
library(haven)

# read files
data_a <- read_sav("data/assignment_4_dataset_A.sav")
data_b <- read_sav("data/assignment_4_dataset_B.sav")

```

## Exploratory data analysis

Run an exploratory data analysis (EDA) to investigate the dataset.

```{r}

# descriptive statistics
summary(data_a)

# check categories 
table(data_a$sex)
table(data_a$pain)

# check distr. of hospitals  
table(data_a$hospital) 
table(data_b$hospital)

```

## Correct coding errors

If you find values in the dataset during the EDA, that are not correct based on the provided descriptions of the variables of the dataset please correct them here.

```{r}

# clean data A
clean_data_a <- data_a %>%
  mutate(sex = tolower(sex)) %>%  
  mutate(sex = ifelse(sex == "woman", "female", sex)) %>%
  mutate(sex = as.factor(sex), hospital = as.factor(hospital)) %>%  
  filter(
    pain >= 0 & pain <= 10,
    mindfulness >= 1 & mindfulness <= 6
  )

# clean data B
clean_data_b <- data_b %>%
  mutate(sex = tolower(sex)) %>%
  mutate(sex = ifelse(sex == "woman", "female", sex)) %>%
  mutate(sex = as.factor(sex), hospital = as.factor(hospital)) %>%  
  filter(
    pain >= 0 & pain <= 10,
    mindfulness >= 1 & mindfulness <= 6
  )

```

## Building the mixed model

First, I fit a linear mixed model on dataset A to predict postoperative pain, using a random intercept for the hospitals

```{r}

#install.packages("lme4")
#install.packages("lmerTest")
#install.packages("MuMIn")

library(lme4)
library(lmerTest)
library(MuMIn)

# fit the random intercept model on dataset A
mixed_model <- lmer(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_saliva + (1 | hospital), data = clean_data_a)

# model coefficients and p-values
summary(mixed_model)

# confidence intervals for the fixed effects
confint(mixed_model)

# compute marginal and conditional R-squared
r.squaredGLMM(mixed_model)

```

## Predicting on a new dataset
Now I use the model trained on dataset A to predict pain in dataset B. Because dataset B contains different hospitals, I only use the fixed effects for prediction

```{r}

# predict pain on dataset B 
predictions_b <- predict(mixed_model, newdata = clean_data_b, re.form = NA)

# Residual Sum of Squares (RSS)
RSS <- sum((clean_data_b$pain - predictions_b)^2)

# Total Sum of Squares (TSS)
TSS <- sum((clean_data_b$pain - mean(clean_data_b$pain))^2)

# R-squared for dataset B
R2_B <- 1 - (RSS / TSS)

cat("The variance explained by the model on datafile B (R-squared) is:", R2_B, "\n")

```

# What to report

The coefficients and 95% confidence intervals for each fixed effect predictor obtained from the linear mixed model on data file A:

| Predictor | Estimate | 95% CI |
| :--- | :--- | :--- |
| (Intercept) | 2.06 | [-0.71, 4.87] |
| Age | -0.03 | [-0.07, 0.01] |
| Sex (Male) | 0.15 | [-0.22, 0.52] |
| STAI Trait | -0.02 | [-0.07, 0.02] |
| Pain Catastrophizing | 0.07 | [0.02, 0.13] |
| Mindfulness | -0.20 | [-0.39, -0.00] |
| Cortisol (Saliva) | 0.66 | [0.43, 0.88] |

As for the variance components, the variance associated with the random intercept (hospital) was 0.37, while the residual variance was 1.45. 

The marginal R^2 of the model on data file A was 0.318 (31.8%), and the conditional R^2 (variance explained by both fixed and random effects) was 0.458 (45.8%). When using this model to predict pain in the new dataset (data file B) using only the fixed effects, the observed R^2 was 0.381 (38.1%).

# What to discuss

Comparing the current linear mixed model with the linear regression model from the previous assignment, several similarities and differences occur. The direction of the relationships between the predictors and postoperative pain remained consistent. Salivary cortisol and pain catastrophizing remained strong positive predictors of pain, whereas mindfulness showed a negative association. 

However, accounting for the clustering of data within different hospitals  likely shifted the exact estimates and widened the confidence intervals compared to the simple linear model. This happens due to the reason that mixed model effectively controls for the shared variance among patients treated at the same site, providing a more conservative and reliable estimate of the fixed effects. 

The variance explained by the fixed effects alone (31.8%) is lower than the total variance explained when including the hospital clusters (45.8%), indicating that institutional differences (e.g., hospital protocols or environments) account for a meaningful portion of postoperative pain variance.

Finally, the cross-validation on data file B yielded an R^2 of 38.1% - it is lower than the conditional R^2 (since we could not use the hospital-specific intercepts for new hospitals), but it is higher than the marginal R^2, suggesting that our fixed psychological and hormonal predictors provide a robust estimation of postoperative pain across completely different hospital settings.